A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-21 text/informal-overview.tex
[WIP] revising overview.

We designed \LMD, a multi-stage calculus with dependent type.
\LMD is based on \LTP\cite{Hanada2014} by Hanada and Igarashi, which is a multi-stage calculus with CSP, and
we introduced dependent types based on \LTP\cite{attapl}.
In this section, we check \LMD informally after checking \LTP and \LLF which are the basis of \LMD.

@@

In \LTP, brackets and escape are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$.
The type of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.
% Please notice that if $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for brackets and escape.
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M 
\end{align*}
It means escape cancels brackets.
% This reduction is called $\longrightarrow_\Lambda$ in \LMD.
% \red{この文はここに来るべきなのか? LTPの説明ではないが}

% stage variable and stage

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage variable}
\footnote{In the original \LTP paper, this is called a \textit{transition variable}. 
However, we call it a \textit{stage variable} in this paper.} and
a sequence of stage variables is called \textit{stage}.
The empty stage is represented by $\varepsilon$.
They are used to show the depth of brackets.
For example, $\TB_\alpha (\lambda x:\I.x+10)$ is the fragment of code which becomes $(\lambda x:\I.x+10)$ after it was run once and
$\TB_\alpha \TB_\beta (\lambda x:\I.x+10)$, which is abbreviated as $\TB_{\alpha\beta} (\lambda x:\I.x+10)$,
becomes $(\lambda x:\I.x+10)$ after it was run it twice.

% stage abstraction and application

There are abstractions for stage variables and applications for stage abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M A$, respectively.
A stage abstraction binds a stage variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
There is a reduction rule for stage application in \LTP.
The rule is the following.
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
For example, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ (\beta\gamma)$ reduces to $\TB_{\beta\gamma} (\lambda x:\I.x)$.

% stage-related symbols disappear when the empty stage is substituted.
% In short, about "run"

Another important rule about a stage variable is 
that symbols with stage variables disappear 
when the empty stage $\varepsilon$ is substituted to the stage variable.
The purpose of this rule is express \verb|run| without any special symbol.
In multi-stage calculus, \verb|run| is a very important operator which changes quoted code to the original code.
For example, $(\TB_\varepsilon (\lambda x:\I.x))$ is equivalent to $(\lambda x:\I.x)$.
Therefore, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ \varepsilon$ becomes $\lambda x:\I.x$
In \LTP, \verb|run| is realized with application to the empty stage $\varepsilon$.

% stage in typing rules

A type judgement of \LTP is of the form $\G \vdash M : \tau @ A$.
A stage in a judgement represents where the judgement is true.
For example, $\G \vdash (\lambda x:\I.x) : \I \to \I @ \alpha$ means 
term $\lambda x:\I.x$ has type $\I \to \I$ at stage $\alpha$.
Especially, terms without quoting exist at the empty stage $\varepsilon$.
For example, $(\lambda x:\I.x)\ (1+2)$ is at $\varepsilon$ stage and 
$\TB_\alpha (\lambda x:\I.x)$ is at $\alpha$ stage.
Terefore, stages appear in typing rules, too.
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A}{\TTB} \andalso
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}}{\TTBL}
\end{center}
\TTB, corresponding to brackets, means 
if $M$ is typed $\tau$ at stage $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
\TTBL\ is converse of \TTB.

@@

CSP, cross-stage persistence, is an important feature of \LTP.
It enables us to embed value at an outer stage into an inner stage.
In \LTP, CSP is written explicitly and $\%$ is a symbol dedicated to CSP
although MetaOCaml CSP is implicit and there is no symbol for CSP.
For example, the variable $x$ is cross-staged into a code value in the following expression.
\[\lambda x:\I.(\Lambda\alpha.(\TB_\alpha (\lambda y:\I.y+\%_\alpha x))\ \varepsilon)\ 10\]

@@

>>>>

We describe our calculus \LMD informally.  \LMD is based on
\LTP~\cite{Hanada2014} by Hanada and Igarashi and we introduce
dependent types based on \LLF~\cite{attapl}.  We start with a review of 
\LTP.

@@

In \LTP, brackets (quasi-quotation) and escape (unquote) are written
$\TB_\alpha M$ and $\TBL_\alpha M$, respectively.  For example,
$\TB_\alpha (1 + 1)$ represents code of expression $1 + 1$ and thus
evaluates to itself.  Escape $\TBL_\alpha M$ may appear under
$\TB_\alpha$; it evaluates $M$ to a code value and splice it into the
surrounding code fragment.  Such splicing is expressed by the
following reduction rule:
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M 
\end{align*}
The type of
$\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$; the type
of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage variable}\footnote{%
  In Hanada and Igarash~\cite{Hanada2014}, it was called a \textit{transition variable}, which is derived from correspondence to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.
Intuitively, they represent the depth of nested brackets.
Stage variables can be abstracted by $\Lambda\alpha.M$ and instantiated by
an application $M\ A$ to stages.
For example, $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10) 5)$ is a code value, where \(\alpha\)
is abstracted.  If it is applied to \(\varepsilon\), which is the empty sequence standing for zero,
\(\TB_\alpha\) becomes \(\TB_\varepsilon\), which stands for zero brackets,
and so \(\TB_\alpha\) disappears.  As a result, the application reduces to (unquoted) \((\lambda x:\I.x+10) 5\) and to 15.  If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
\(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).  These are expressed by
the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB\) and \(\TBL\) (and also \(\%_\alpha\) as we see later).  As we
show above, application to $\varepsilon$ corresponds to \texttt{run}.

Cross-stage persistence (CSP), which is an important feature of \LTP,
is a primitive to embed values (not necessarily code values) into a
code value.
For example,
\begin{verbatim}
   let y = 1 + 2 in
   .< y * 2 >.    (* .< e >. stands for brackets *)
\end{verbatim}
evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
Here, CSP is implicitly applied to the occurrence of \texttt{y}
and embeds its value \texttt{3} into the code.
In \LTP, CSP is written explicitly and $\%$ is a symbol for CSP.
For example, the above MetaOCaml expression can be expressed
in \LTP as follows:
\[
  (\lambda x:\I.\Lambda\alpha.(\TB_\alpha ((\%_\alpha y) * 2)))\ (1 + 2)
\]
It evaluates to \(\Lambda\alpha.(\TB_\alpha ((\%_\alpha 3) * 2))\), in
which $\%_\alpha 3$ means that it waits for the surrounding code to be
run (by an application to $\varepsilon$).  If it is applied to
\(\varepsilon\), substitution of \(\varepsilon\) for \(\alpha\)
eliminates \(\TB_\alpha\) and \(\%_\alpha\) and so \(3 * 2\) is
obtained.

The type system of \LTP uses code types---the type of code of type
\(\tau\) is written \(\TW_\alpha \tau\)---for typing \(\TB_\alpha\),
\(\TBL_\alpha\) and \(\%_\alpha\).  It takes stages into account: A
variable declaration in a type environment is associated with its
declared stage as well as its type and a type judgement of \LTP is of
the form $\G \vdash M : \tau @ A$, in which $A$ stands for the stage
of term $M$.  For example,
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \alpha$ holds, but
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \varepsilon$ does
not because the latter uses $y$ at stage \(\varepsilon\) but $y$ is
declared at $\alpha$.  Quotation \(\TB_\alpha M\) is given type
\(\TW_\alpha \tau\) at stage \(A\) if \(M\) is given type \(\tau\) at
stage \(A\alpha\) and  unquote \(\TBL_\alpha M\) is given type
\(\tau\) at stage \(A\alpha\) if \(M\) is given type \(\TW_\alpha \tau\) at
stage \(A\alpha\), represented by the following rules.
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A}{\TTB} \andalso
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}}{\TTBL}
\end{center}
% \TTB, corresponding to brackets, means 
% if $M$ is typed $\tau$ at stage $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
% \TTBL\ is converse of \TTB.

@@

@@

\AI{We should discuss conceptual points, not technical points!}
